Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to agroecossystems

### 2021-08-04 Manuscript/Manuscript.tex
jh

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States . The species has showed a remarkable
capacity to evolve resistance to herbicides. To date, Palmer amaranth
has evolved resistance to eight herbicide sites of action (Heap, 2021),
increasing the weed management complexity (Lindsay et al., 2017).
Unmanaged Palmer amaranth in competition for water, light and nutrients
can drastically impact on crop yields (Berger et al., 2015). For
example, Palmer amaranth is documented with potential to reduce 91\%,
68\%, and 54\% of corn (Massinga et al., 2001), soybean (Klingaman and
Oliver, 1994), and cotton (Morgan et al., 2001) yields, respectively.
Thus, Palmer amaranth poses an economical risk to sustainable
agriculture.

@@

Light and temperature are likely the main environment requirements for
Palmer amaranth successful adaptation. Palmer amaranth is reported with
an extended germination period (Jha et al., 2010). Germination of Palmer
amaranth is triggered by 18 C soil temperature (Keeley et al., 1987),
and optimal germination and biomass production occur at 35/30 C day and
night temperatures (Guo and Al-Khatib, 2003). Water has not shown to
limit Palmer amaranth fitness. Under continuous water stress, Palmer
amaranth survived and produced at least 14000 seeds plant-1 (Chahal et
al., 2018). Also, seeds from Palmer amaranth growing under water stress
conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). The continuous global temperature warming can
impact agriculture and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already documented in agronomic crops of South
America (Larran et al., 2017; Küpper et al., 2017) and Southern Europe
(Milani et al., 2021). In the US, Palmer amaranth is well established at
crop (Garetson et al., 2019) and non-crop land (Bagavathiannan and
Norsworthy, 2016) in the warm southern United States but its range is
expanding to cool temperatures northward. For example, herbicide
resistant Palmer amaranth is widespread in Nebraska (Oliveira et al.,
2021), Michigan (Kohrt et al., 2017), and Connecticut (Aulakh et al.,
2021). Successful cases of Palmer amaranth invasion and near to
eradication is reported in Minnesota (Yu et al., 2021). No Palmer
amaranth actively growing was found in Canada; however, Palmer amaranth
seeds was detected in sweet potato slips (Page et al., 2021).
Nonetheless, it seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Therefore, strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Strategies to manage Palmer amaranth should
be built on minimizing the species ability to adapt, grow and develop
into agroecossystems.

The continuous Palmer amaranth dispersal and potential establishment
into northern United States warrant investigations on species morphology
in such environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under corn, soybean and fallow at two timings
across five locations in the United States Midwest.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to fallow and soybean (Figure 2A). Palmer amaranth reached 10\%
flowering in soybean at doy 180, which was slightly different from
fallow (doy 180.9; \emph{P} = 0.01) and corn (doy 181.7; \emph{P} =
0.00). Nonetheless, the 10\% cumulative Palmer amaranth flowering in
soybean, fallow and corn occurred at the end of June. The 50\% Palmer
amaranth cumulative flowering occurred in July. For example, Palmer
amaranth reached 50\% in fallow at doy 193.4, followed by soybean (doy
194.8), corn (doy 206.6). Similar trend was observed at 90\% Palmer
amaranth cumulative flowering. Palmer amaranth growing in corn reached
90\% flowering at doy 252.6 (early September), which was 37.8 and 32.2
days after Palmer amaranth 90\% flowering in fallow and soybean,
respectively.

Palmer amaranth cumulative flowering from second transplanting timing
ranged from mid July to mid September (Figure 2). All cumulative
flowering comparisons among agroecossystems were significant (\emph{P}
\textless{} 0.00) Palmer amaranth growing in fallow resulted in earlier
flowering time compared to soybean and corn. Palmer amaranth growing in
fallow reached 10\%, 50\%, and 90\% flowering time at day 203.8, 214.4,
and 232.2, respectively. Palmer amaranth growing in soybean reached 10\%
flowering at doy 210.9, which was 6 days prior to corn (\emph{P}-value =
0.00). Similar trend was observed at 50\% flowering, whereas Palmer
amaranth reached 50\% flowering in corn (doy 233.0) 4 days after
soybeans (doy 228.9; \emph{P} = 0.00). The 90\% Palmer amaranth
cumulative flowering occurred at same day in corn (260.9) and soybean
(260.5; \emph{P} = 0.66).

@@

Palmer amaranth accumulated more biomass when growing in fallow compared
to Palmer amaranth growing in soybean and corn (Figure 3). At first
transplanting time, Palmer amaranth biomass was 78.3 g plant-1, 26.4 g
plant-1 and 14.9 g plant-1 in fallow, soybean and corn, respectively.
Palmer amaranth produce less biomass at second transplanting time. For
example, Palmer amaranth growing in fallow resulted in 49.2 g plant-1,
3.4 g plant-1 in soybean and 1.4 g plant when growing in corn.

For example, Palmer amaranth growing at first cohort accumulated 90\%,
85\% and 31\% more biomass when compared to second cohort at corn,
soybean and fallow, respectively

In addition, Palmer amaranth growing in corn and soybean at second
cohort had a similar end-season flowering pattern.

@@

Our study showed the Palmer amaranth flowering pattern, grow and
development varied within agroecossystems and timing. In general, Palmer
amaranth produced more biomass and taller plants when growing at first
cohort rather than second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
both species, crop and weed. Thus, Palmer amaranth growing strategy was
to mimic the crop species (Figure 4). In contrast, at second cohort,
Palmer amaranth grow and development was limited due to the crop
competitive ability at advanced development stages.

Taller Palmer amaranth plants is a strategy to compete for light in
between crop rows in absence of canopy. For example, at first cohort,
corn and Palmer amaranth had abundant light to grow and develop, which
likely aid its extended the flowering pattern. However, at second
cohort, Palmer amaranth was transplanted when corn canopy was nearly
closed, which reduced Palmer amaranth competition. As a result, Palmer
amaranth height and biomass was nearly 50\% and 90\% lower compared to
its first cohort, respectively. Similar trend was observed to Palmer
amaranth growing in soybeans. Palmer amaranth growing without crop
competition produced highest amount of biomass. The Palmer amaranth
strategy in fallow was to invest biomass in growing plant width and
height. Nonetheless, Palmer amaranth produced 31\% less biomass in
second cohort compared to first cohort timing. Therefore, crop
competition was not the only factor limiting Palmer amaranth growing in
the second cohort.

It was not evaluate seed production due to Palmer amaranth harvest at
early flowering. Nonetheless, it is reported a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated nearly 50\% more overall
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the agroecossystem. Seed production and deposition in the
seedbank is a key factor for species perpetuation (Menges, 1987). Palmer
amaranth germination is influenced by soil depth and seed viable seeds
were fount after 36 months buried in the seedbank (Sosnoskie et al.,
2013). Therefore, tactics that promote crop advantage against Palmer
amaranth, including early crop planting, preemergence applied herbicide,
and crop residue (e.g.~cover crops) can minimize the negative impact of
Palmer amaranth in agroecossystems.

Palmer amaranth has shown to be responsive to to thermal units (e.g,
growing degree days). A study in North Carolina demonstrated that Palmer
amaranth emergence in response to soil temperature, with seedlings
emerging before July prior to crop canopy closure (Piskackova et al.,
2021). In addition, in a bareground study, Palmer emergence emergence
pattern in Nebraska was influenced by soil management. Early tillage
resulted in 90\% Palmer amaranth emergence in June

can cause significant soybean yield loss (Korres et al., 2019)

@@

@@

@@

@@

\leavevmode\hypertarget{ref-korres2019}{}%
Korres, N. E., Norsworthy, J. K., and Mauromoustakos, A. (2019). Effects
of {Palmer Amaranth} ({Amaranthus} palmeri) {Establishment Time} and
{Distance} from the {Crop Row} on {Biological} and {Phenological
Characteristics} of the {Weed}: {Implications} on {Soybean Yield}.
\emph{Weed Science} 67, 126--135.
doi:\href{https://doi.org/10.1017/wsc.2018.84}{10.1017/wsc.2018.84}.

@@

@@

@@

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States. Unmanaged Palmer amaranth in competition
for water, light and nutrients can drastically impact on crop yields
(Berger et al., 2015). For example, Palmer amaranth is documented with
potential to reduce 91\%, 68\%, and 54\% of corn (Massinga et al.,
2001), soybean (Klingaman and Oliver, 1994), and cotton (Morgan et al.,
2001) yields, respectively. Moreover, Palmer amaranth has showed a
remarkable capacity to evolve resistance to herbicides. To date, Palmer
amaranth has evolved resistance to eight herbicide sites of action
(Heap, 2021), increasing the weed management complexity (Lindsay et al.,
2017). Thus, Palmer amaranth poses an economical risk to sustainable
agriculture.

@@

Light and temperature are likely the main environment requirements for
Palmer amaranth successful grow and development. Palmer amaranth is
reported with an extended germination period (Jha et al., 2010).
Germination of Palmer amaranth is triggered by 18 C soil temperature
(Keeley et al., 1987), and optimal germination and biomass production
occur at 35/30 C day and night temperatures (Guo and Al-Khatib, 2003).
Also, Palmer amaranth germination are human-mediated by tillage timings
and preemergence-applied herbicides (Chahal et al., 2021). Water has not
shown to limit Palmer amaranth fitness. Under continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds plant-1
(Chahal et al., 2018). Also, seeds from Palmer amaranth growing with
limited water conditions were heavier, less dormant, and prompt for
germination (Matzrafi et al., 2021). The continuous global temperature
warming can impact agriculture and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is well established at cotton
fields (Garetson et al., 2019) and non-crop areas (Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021), Michigan (Kohrt et al.,
2017), Connecticut (Aulakh et al., 2021), and potential damaged to
soybean fields in Illinois is reported (Davis et al., 2015). Successful
cases of Palmer amaranth invasion and near to eradication is documented
in Minnesota (Yu et al., 2021). No Palmer amaranth actively growing was
found in Canada; however, Palmer amaranth seeds was detected in sweet
potato slips (Page et al., 2021). Nonetheless, it seems fated the need
to manage Palmer amaranth in agronomic crops throughout multiple
environments in the near future. Strategies on Palmer amaranth
management should encompass the agroecosystem level but not only
attempts to eradicate the weed. Most tactics to manage Palmer amaranth
are based on technology fixes (Scott, 2011), which are short-term (e.g.,
herbicide and/or tillage) rather than long-term weed management. Palmer
amaranth management should be built on minimizing the species ability to
adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed (Yu
et al., 2021). Nonetheless, the continuous Palmer amaranth dispersal and
potential establishment across northern United States is concerning and
warrant investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecossystems can enhance our knowledge on species
adaptation. It can also aid on designing proactive and ecological
tactics to limit the species range expansion, reduce its negative
impact, and design resilient and sustainable farming systems (MacLaren
et al., 2020). Therefore, the objective of this study was to investigate
the flowering pattern, biomass production, and height of Palmer amaranth
growing under corn, soybean and fallow at two timings across five
locations in the United States Midwest.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to fallow and soybean at first cohort (Figure 2A). The 10\%
cumulative Palmer amaranth flowering in soybean, fallow and corn
occurred at the end of June. Palmer amaranth reached 10\% flowering in
soybean at doy 180, which was slightly different from fallow (doy 180.9;
\emph{P} = 0.01) and corn (doy 181.7; \emph{P} = 0.00). The 50\% Palmer
amaranth cumulative flowering occurred in July. For example, Palmer
amaranth reached 50\% in fallow at doy 193.4, followed by soybean (doy
194.8), corn (doy 206.6). Similar trend was observed at 90\% Palmer
amaranth cumulative flowering. Palmer amaranth growing in corn reached
90\% flowering at doy 252.6 (early September), which was 37.8 and 32.2
days after Palmer amaranth 90\% flowering in fallow and soybean,
respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (Figure 2). All cumulative flowering comparisons
among agroecossystems were significant different (\emph{P} \textless{}
0.00) Palmer amaranth growing in fallow resulted in earlier flowering
time compared to soybean and corn. Palmer amaranth growing in fallow
reached 10\%, 50\%, and 90\% flowering time at day 203.8, 214.4, and
232.2, respectively. Palmer amaranth growing in soybean reached 10\%
flowering at doy 210.9, which was 6 days prior to corn (\emph{P}-value =
0.00). Similar trend was observed at 50\% flowering, whereas Palmer
amaranth reached 50\% flowering in corn (doy 233.0) 4 days after
soybeans (doy 228.9; \emph{P} = 0.00). The 90\% Palmer amaranth
cumulative flowering occurred at same day in corn (260.9) and soybean
(260.5; \emph{P} = 0.66).

@@

Palmer amaranth accumulated more biomass when growing in fallow compared
to Palmer amaranth growing in soybean and corn (Figure 3A). At first
cohort time, Palmer amaranth biomass was 78.3 g plant-1, 26.4 g plant-1
and 14.9 g plant-1 in fallow, soybean and corn, respectively. Moreover,
Palmer amaranth produce less biomass at cohort time. Palmer amaranth
growing in fallow resulted in 49.2 g plant-1, 3.4 g plant-1 in soybean
and 1.4 g plant when growing in corn.

The Palmer amaranth achieve 92.3 cm tall when growing at first cohort
timing in corn (Figure 3B).

@@

Our study showed that Palmer amaranth flowering time, biomass and height
varied within agroecossystems and cohort timing. In general, Palmer
amaranth produced more biomass and taller plants when growing at first
cohort rather than second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
both species, crop and weed. High biomass and taller Palmer amaranth
plants is likely a weed strategy to compete for light in between crop
rows in absence of canopy. In such condition, Palmer amaranth showed an
extraordinary plasticity to adapt upon the agroecosystem. This is
evident when comparing Palmer amaranth shape, and its extended flowering
pattern when growing into corn compared to soybean. In corn, Palmer
amaranth was taller and thinner when growing in corn. The Palmer
amaranth competition strategy was likely to mimic the crop grow and
development (Figure 4). These results are concerning as breeding more
competitive crop varieties is likely to select more competitive Palmer
amaranth biotypes (Bravo et al., 2017).

Palmer amaranth grow and development at second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy, Palmer amaranth flowering pattern was near to similar into corn
and soybean. Palmer amaranth growing without crop competition produced
highest amounts of biomass and lowest extended flowering pattern. The
Palmer amaranth strategy in bareground was to invest biomass in growing
plant width and height. Nonetheless, Palmer amaranth produced 31\% less
biomass in second cohort compared to first cohort timing. In a
bareground study, early emerged Palmer amaranth without competition was
50\% taller than late emerged plants (Webster and Grey, 2015). These
results suggests that crop competition is not the only factor limiting
late Palmer amaranth establishment. The limited growth of Palmer
amaranth at second cohort is likely a reduced plant response to thermal
units (e.g, growing degree days). It is hypothesize that reduced day
length contributed to smaller plants at second cohort as well as shorter
flowering period. A study in North Carolina and Illinois predicted that
less than 10\% Palmer amaranth seedlings emergence occurred after June
(Piskackova et al., 2021). In addition, Palmer amaranth negative impact
on soybean (Korres et al., 2020) and cotton (Webster and Grey, 2015)
yields was higher when plants were established near to crop planting.

Seed production was not evaluated due to Palmer amaranth harvest at
flowering time. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated an overall 50\% more
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the agroecossystem. Our observation is consistent with the
findings that first Palmer amaranth cohort produced 50\% more seeds per
plant than Palmer amaranth plants established six weeks later in
bareground (Webster and Grey, 2015). Nonetheless, seed production at
second cohort is likely to replenish the soil seedbank. Seed production
and deposition in the seedbank is a key factor for species perpetuation
(Menges, 1987). Palmer amaranth can produce hundred thousands seeds per
plant (Schwartz et al., 2016; Keeley et al., 1987), and stay viable
buried in the seedbank for at least 36 months (Sosnoskie et al., 2013).

Our study demonstrated the short-term Palmer amaranth plasticity to grow
and develop into cropping-systems. Palmer amaranth management priority
should focus on minimizing Palmer amaranth dispersal into new
geographies. Early-season management programs would have a large
negative effect in Palmer amaranth growth and development. Tactics that
promote crop advantage against Palmer amaranth, including early crop
planting, crop rotation, plant width, preemergence applied herbicide
(Sanctis et al., 2021), and crop residue (e.g.~cover crops) would
minimize the negative impact of Palmer amaranth in agroecossystems.

@@

\leavevmode\hypertarget{ref-bravo2017}{}%
Bravo, W., Leon, R. G., Ferrell, J. A., Mulvaney, M. J., and Wood, C. W.
(2017). Differentiation of {Life}-{History Traits} among {Palmer
Amaranth Populations} ({Amaranthus} palmeri) and {Its Relation} to
{Cropping Systems} and {Glyphosate Sensitivity}. \emph{Weed Science} 65,
339--349.
doi:\href{https://doi.org/10.1017/wsc.2017.14}{10.1017/wsc.2017.14}.

@@

\leavevmode\hypertarget{ref-chahal2021}{}%
Chahal, P. S., Barnes, E. R., and Jhala, A. J. (2021). Emergence pattern
of {Palmer} amaranth ({Amaranthus} palmeri) influenced by tillage
timings and residual herbicides. \emph{Weed Technology} 35, 433--439.
doi:\href{https://doi.org/10.1017/wet.2020.136}{10.1017/wet.2020.136}.

@@

\leavevmode\hypertarget{ref-davis2015}{}%
Davis, A. S., Schutte, B. J., Hager, A. G., and Young, B. G. (2015).
Palmer {Amaranth} ({Amaranthus} palmeri) {Damage Niche} in {Illinois
Soybean Is Seed Limited}. \emph{Weed Science} 63, 658--668.
doi:\href{https://doi.org/10.1614/WS-D-14-00177.1}{10.1614/WS-D-14-00177.1}.

@@

\leavevmode\hypertarget{ref-korres2020}{}%
Korres, N. E., Norsworthy, J. K., Mauromoustakos, A., and Williams, M.
M. (2020). Soybean density and {Palmer} amaranth ({Amaranthus} palmeri)
establishment time: Effects on weed biology, crop yield, and economic
returns. \emph{Weed Science} 68, 467--475.
doi:\href{https://doi.org/10.1017/wsc.2020.41}{10.1017/wsc.2020.41}.

@@

\leavevmode\hypertarget{ref-macrae2013}{}%
MacRae, A. W., Webster, T. M., Sosnoskie, L. M., Culpepper, A. S., and
Kichler, J. M. (2013). Cotton {Yield Loss Potential} in {Response} to
{Length} of {Palmer Amaranth} ({Amaranthus} palmeri) {Interference}. 17,
6.

@@

\leavevmode\hypertarget{ref-sanctis2021}{}%
Sanctis, J. H. S. de, Barnes, E. R., Knezevic, S. Z., Kumar, V., and
Jhala, A. J. (2021). Residual herbicides affect critical time of
{Palmer} amaranth removal in soybean. \emph{Agronomy Journal} 113,
1920--1933.
doi:\href{https://doi.org/10.1002/agj2.20615}{10.1002/agj2.20615}.

@@

\leavevmode\hypertarget{ref-webster2015}{}%
Webster, T. M., and Grey, T. L. (2015). Glyphosate-{Resistant Palmer
Amaranth} ({Amaranthus} palmeri) {Morphology}, {Growth}, and {Seed
Production} in {Georgia}. \emph{Weed Science} 63, 264--272.
doi:\href{https://doi.org/10.1614/WS-D-14-00051.1}{10.1614/WS-D-14-00051.1}.
